---
title: "week16_recap"
author: "Daniel Trielli"
date: "2023-12-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## Loading the packages

Run the codeblock below to load the packages we will need for this recap

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(tigris)
library(tidycensus)
library(corrr)
library(tidytext)
```

## Load Data

[USAFacts compiled data on inspections on child care facilities throughout the country](https://usafacts.org/data-projects/childcare-inspections) - specifically, how many of those facilities are overdue for inspection. Let's take a look on the data they have on Maryland.

First, here the data we need to load:

```{r}
childcare_inspections <- read_csv('data/childcare_inspections.csv')
childcare_inspections$fips <- as.character(childcare_inspections$fips)
```

#### **Q1** The childcare data does not have the names of the counties. Join them with the md_counties data, that can be loaded in the block below. Then inspect the dataframe you created with this join and answer this: there are two columns with county names in the new dataframe. Which one is the best to be used? (hint: it has to do with homonyms)


```{r}
# This loads the county data
md_counties <- counties() |>  filter(STATEFP == "24") |>
  clean_names()

```

```{r}
# Join the childcare data with the county data 

childcare_and_counties <- childcare_inspections |>
 inner_join(md_counties, by=c("fips"= "geoid"))

```

##### A1: The namelsad column is the best to be used, because it distinguishes Baltimore City and Baltimore County. The other column just has "Baltimore," which could lead to data being joined the wrong way.

------------------------------------------------------------------------

#### **Q2** Create bar chart with the percentages of overdue childcare facility inspections per county to answer (You'll need to create a column for that calculation first). What is the county with the largest percentage of facilities with overdue inspections?

```{r}
# Calculate the percentage of overdue inspections per county
childcare_and_counties <- childcare_and_counties |>
  mutate(overdue_inspections =(overdue/total)*100) |>
  arrange(desc(overdue_inspections))

# Create the chart with ggplot
 ggplot(childcare_and_counties, aes(x=namelsad, y=overdue_inspections)) +
  geom_bar(stat = "identity") +
   

```

##### **A2:** Montgomery County has the largest percentage of facilities with overdue inspections

------------------------------------------------------------------------

#### **Q3** Next, we want to see if the number of child care facilities - with overdue inspections or otherwise - is appropriate for each county. So we're going to calculate the rate of children per facility per county. To do that, you'll first need Census data to get the population of children under 5 for each county. Which county has the biggest rate of children per facility (i.e. they could use more facilities)?

```{r}
# First, grab data from the ACS. The ACS variable for children age under 5 is B06001_002 We want 2022 data.
child_population <- get_acs(geography = "county", 
          variables = c(under_five ="B06001_002"), 
          state="MD", 
          year= 2022)

```

```{r}
# Rename the column name 'estimate', which has our population data, with something more useful, like 'children_pop'
child_population <- child_population |>
  clean_names() |>
  rename(children_pop = estimate)

```

```{r}
# Join the dataframes
childcare_with_population <- childcare_and_counties |>
 inner_join(child_population, by=c("fips"= "geoid"))

# Calculate the rates for children per facility
childcare_with_population <- childcare_with_population |>
  mutate(children_per_facility = children_pop/total) |>
  arrange(desc(children_per_facility))

```

##### **A3:** Garrett County has the highest rate of children per facility.

------------------------------------------------------------------------

#### **Q4** Make a map of the results for question 3. Are there any areas of the state that are more lacking in childcare facilities?

```{r}
ggplot() +
  geom_sf(data=childcare_with_population, aes(fill=children_per_facility, geometry=geometry)) +
  scale_fill_viridis_c(option="magma", trans="log")

```

##### **A4**: Southern and Eastern Maryland are lacking in childcare facilities and have a much higher rate of children per facility than other parts of the state.

------------------------------------------------------------------------

#### **Q5** Now let's explore if we can find patterns in under-served areas, particularly when it comes to median income of the population. First, join the childcare data median income data from the Census. Then, create a scatterplot and perform a cor.test(), and answer: is there is any correlation between income and rate of children per facilities? What makes you conclude that?

```{r}
# Again, grab data from the ACS. The ACS variable for median income is B19013_001. We want 2022 data again.

median_income <- get_acs(geography = "county", 
          variables = c(median_income ="B19013_001"), 
          state="MD", 
          year= 2022)
```

```{r}
# Rename the column name 'estimate', which has our income data, with something more useful, like 'median_income'

median_income <- median_income |>
  clean_names() |>
  rename (median_income = estimate)

```

```{r}
# Join the dataframes
income_childcare <- childcare_with_population |>
 inner_join(median_income, by=c("fips"= "geoid"))

```

```{r}
# Create a scatterplot with median income and rate of children per facility
income_childcare |>
ggplot() +
  geom_point(aes(x=median_income,y=children_per_facility)) +
  geom_smooth(aes(x=median_income,y=children_per_facility), method="lm")

```

```{r}
# Create a correlation test.
cor.test(income_childcare$median_income, income_childcare$children_per_facility)

```

##### A5: There is not much of a correlation here. The correlation is -0.11, which demonstrates a very slight relationship between counties with higher median incomes and counties with lower rates of children per childcare facility (which indicates less crowded childcare facilities). But this number is very small, and as we can see on the scatterplot, there isn't a very evident relationship. Also, the p-value is 0.61, which indicates a high chance any relationship between those two variables is just by chance and doesn't actually mean anything.

------------------------------------------------------------------------

#### **Q6** Finally, let's do some text analysis. We another dataset, with inspection reports from informal child care providers in the state. This dataset has an informative column of initial findings: a column that describes issues found in these providers, if any. Your goal is to create bigrams for this column and find common issues that are mentioned in these findings. And then answer: what kind of issues listed there you think are interesting? They don't have the most frequent.

```{r}
# Read the reports data
reports <- read_csv('data/childcare-informal-provider-inspections.csv') |> clean_names()
```

```{r}
# Create a stopword dictionary (feel free to include more words)
unique_words <- reports |> select(initial_findings) |>
  unnest_tokens(word, initial_findings)

view(unique_words)


```

```{r}
data("stop_words") 

stop_words <- stop_words |>
  add_row(word = "compliant") |>
  add_row (word = "with") |>
  add_row (word = "h") |>
  add_row (word = "s") |>
  add_row (word = "standards")
```


```{r}
# Unnest tokens into bigrams, removing stop words and repasting bigrams, list bigrams


common_bigrams <- reports |>
  unnest_tokens(bigram, initial_findings, token = "ngrams", n = 2) |>
  separate(bigram, c("word1", "word2"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |>
  mutate(bigram = paste(word1, word2, sep=" ")) |>
  group_by(bigram) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(20)

```

##### A6: I think that "knife draw" and "knives moved" indicate the most interesting issues, because they likely are referring to improper storage of dangerous household objects, which seems like a frequent issue that childcare providers are cited on. "Missing ecma," "ecma docs," "emergency docs," and several similar terms also pop up. These are interesting because they point to crucial documents being missing from childcare providers' facilities, which leads us to wonder what those documents are, and why so many providers didn't have them when inspected.