---
title: "data_analysis_basketball"
output: html_document
date: "2024-04-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(janitor)
library(lubridate)
library(tidyverse)
library(tidycensus)
census_api_key("ef5457f8a2b11d9eecec191d6867f76f96db192b")
```
```{r}
roster_23_24 <- read_csv("data/wbb_rosters_2023_24.csv")
roster_22_23 <- read_csv("data/wbb_rosters_2022_23.csv")
roster_21_22 <- read_csv("data/wbb_rosters_2021_22.csv")
roster_20_21 <- read_csv("data/wbb_rosters_2020_21.csv")
```
```{r}
#Appending rosters for all four seasons 
all_rosters <- bind_rows(roster_20_21, roster_21_22, roster_22_23, roster_23_24)

```

```{r}
#Changing column types to character as needed

all_rosters <- all_rosters |>
mutate(
  ncaa_id = as.character(ncaa_id), 
  player_id = as.character(player_id)
)
```


Data limitations
- The four rosters are inconsistent in the number of columns they have — we think this probably won't matter, because all the columns that we specifically are working with are present in all the datasets, but that is something to consider limitation-wise
- For some players who are graduate students, the data lists their former (undergraduate) college under the high school column. This is just a quirk of how the data is recorded, and won't really impact us — because it's such a rare case, these colleges will likely show up at the bottom of the list when we're analyzing what high schools produced the most players, and thus won't factor into our most important results. But it's something to keep an eye out for.
-Some players here and there are missing certain pieces of information. Most instances make sense — for instance, high school is mostly not listed for international players. Other pieces of information like primary position are also missing, but those shouldn't impact our analysis because we're not looking at those columns. But there are also some players that are missing more important recorded information like heights — when analyzing height in question 5, how should we go about this? We could just drop those players from the analysis and note that we did that in a written explanation in our document. 
-As shown in our process for solving question 2, the way that high schools are recorded is inconsistent. Some high schools in different states with the same name show up stylized the same way in the hs_clean column, and some high schools that are the same showup differently in that column (e.g. Centennial vs. Centennial HS). This requires us to be creative and do some approximation and manual checking when figuring out the answer to question 2, which we explain in code comments under that question. This also means that the result for that question may not be 100% accurate, and rather, is an approximation using the constraints we're given.


Timeframe
-Timeframe should be fine -- for most of the questions we're investigating, we're going to only use the 2023-24 dataset, and for one question that requires comparison across years, we have those datasets present. We'll just specify in our project that we used the timeframe of data available to us, which encompasses the 2020-2024 seasons. 
-Some data is slightly outdated. For instance, the PAC-12 conference listed in some of the rosters does not exist anymore. But since we're not analyzing anything by conference, this should also be fine

Missing definitions/context
- Things we need definitions/data dictionary for: count, redshirt (see bullet below for specifics)
-For the redshirt column, it would be good to get clarification on how that was documented in the data. Did the players marked as having redshirted redshirt in the season that the dataset represents, or does that just indicate that they redshirted at any point in their personal college career?

What we did to make data better-defined
- Appended rosters across all four years into a single dataset. This will make it easier to see comparisons of international players across years to help us answer question 3
- Changed column types as needed to convert things to characters (e.g. player ID) that were registering as numbers but actually are not

```{r}
#Basic analysis to show number of rows and columns in each dataset
roster_20_21

roster_21_22

roster_22_23

roster_23_24

```


Basic exploratory analysis shows that...
- 2020-21 roster has 29 columns and 13075 rows (players)
- 2021-22 roster has 26 columns and 13645 rows (players)
- 2022-23 roster has 29 columns and 13805 rows (players)
- 2023-24 roster has 31 columns and 13907 rows (players)
- When all the rosters are combined, the appended dataset shows that there have been 54,432 players across the past four seasons


1. What states produced the most NCAA players for the 2023-2024 season? What about states that produced the most players per capita? 

Why we're interested: We want to look at this question to see if there are trends in higher-performing states athletically in the U.S. — it would be interesting to form hypotheses about what factors might drive some states to produce more athletes than others. 
```{r}
roster_23_24 |>
#For players who come from places outside the U.S., state_clean is listed as NA (because they don't have a state). This line of code will drop all those instances since we are only looking at American players for this question 
  drop_na(state_clean) |>
  group_by(state_clean)|>
  summarise (
    state_count = n()
  ) |>
  arrange(desc(state_count)) 
#We can see here that Texas produced the most players, followed by Pennsylvania and New York. Guam, the Virgin Islands and Puerto Rico produced the fewest players, which is also expected given that those are small territories.  
```

```{r}
#But looking at numbers alone may not give us a great sense of things. For instance, Texas is one of the more populated states, so it makes sense that it would produce more players. Instead, we're going to try and look at the per capita rate, by attaching a dataset with the population of each state. We call this dataset by using the Census API. This will give us a better sense of the states that produced the most players relative to their size and population.

#This is us pulling state abbreviations from the Census data so we can join that, which will then make it easier to join our population data (which we get in the next step) to the roster data 
data(fips_codes) 

state_abbrev <- fips_codes |> 
  select(state,state_name,state_code) |> 
  distinct()

state_population <- get_acs(geography = "state", 
        variables = c(pop18plus = "B09021_001"), 
        year= 2019) |>
  clean_names() |>
  rename(state_name = name)

#Now we're going to join these so we can see population based on state abbreviation, rather than full state name 
state_abbrev_population <- state_abbrev|>
 inner_join(state_population, by=c("state_name"))

#Now, we're joining this new table to the 2023-24 roster so we can attach populations to the number of players produced in each state. The last line of code also removes all players not from a state (international players), because we don't care about those numbers in this question
player_numbers_by_state <- roster_23_24 |>
  group_by(state_clean) |>
  summarise (
    player_count = n()
  ) |>
  filter(!is.na(state_clean))

state_players_and_population <- player_numbers_by_state |>
  left_join(state_abbrev_population, by=c("state_clean" = "state")) |>
  rename(population = estimate) |>
  filter(!is.na(state_name))

#Finally, we create a new column and calculate the per capita rate of players in each state per 100,000 people 
state_players_and_population |>
  mutate(per_capita_rate = (player_count/population) *100,000) |>
  arrange(desc(per_capita_rate))

```

Answer: The top five states to produce players numbers-wise are Texas, Pennsylvania, New York, California, Ohio and Illinois, which is expected given that those are some of the ore populated states. For further analysis, when we look at the rates of players per 100,000 residents, the top states are Minnesota, Vermont, South Dakota, Wisconsin and Pennsylvania. It's interesting that there's not much overlap in these two measures — although states like Texas and New York may produce a lot of players, they don't produce an above-average amount given their populations. But seeing that smaller states like Vermont and South Dakota produce a lot of players indicates that despite a lack of population, these states are exceptional in sending women to NCAA basketball teams.


2. What high schools produced the most NCAA recruits for the 2023-24 season? Are they mostly public or private?

Why we're interested: We're looking at this question to see if there are any exceptional high schools in the United States. We're also curious about if the top athlete-producing schools tend to lean public or private — that might raise some indications about how socioeconomic status and available educational resources may influence the likelihood of a player going to the NCAA. 

```{r}
roster_23_24 |>
#For players who come from places outside the U.S., hs_clean often isn't listed. We're going to drop these instances since we're only interested in the high schools that are included in the data 
  drop_na(hs_clean)|>
  group_by(hs_clean)|>
  summarise(
    high_schools = n()
  ) |>
  arrange(desc(high_schools)) |>
#For the scope of our project, let's just look at the top 10 high schools. This line of code actually gives us 12 high schools, but that's because there are some that produced the same amount of players as others.
  top_n(10) 
```

```{r}
#We run into an issue here where some high schools from different states are stylized the same way in the hs_clean column, conflating different high schools in the same count. So we're going to create a new column distinguishing these by high school name AND state instead, and then group by that to get a more accurate result. 

roster_23_24 |>
  drop_na(hs_clean)|>
  mutate(hs_and_state = paste(hs_clean, "-", hometown_clean)) |>
  group_by(hs_and_state)|>
  summarise (
    hs_count = n()
  ) |>
  arrange(desc(hs_count)) |>
  top_n(10)

#After examining these results, we see another, different issue. While this gets rid of the multiple Centennial high schools in different states being lumped together, for instance, it poses a different problem. It omits certain schools that seem to be boarding schools, because all the players from those schools have different hometowns listed. But schools like New Hope and IMG Academy are some of the top contenders. To reconcile these issues, we opted to hand-compare the top 10 results from each of the blocks of code, to compile our own list of the actual top 10 schools. The list is below this codeblock. One limitation: these may not be entirely accurate. Based on this data alone (since people may have a different hometown than where they went to high school), it's hard to tell what high schools should be taken out, because it's unclear if there are multiple high schools with the same name. But this is an approximation based on our best guesses. 
```
Answer: The top 10 high schools that produced players are: New Hope Academy, IMG Academy, Notre Dame Academy, Sacred Heart Academy, Trinity, Example Academy, DeSoto High School, Dunmore, South Mecklenburg, and Centennial (in Las Vegas).

After looking at the result from this code and conducting outside research, we can see that 6  are private schools and 4  are public schools. The top 3 high schools that produce players are all private schools, showing a trend that private schools tend to produce more NCAA players than public schools. 

3. Are there changes between school years in terms of how many student athletes outside of the U.S. are on the NCAA roster? 

Why we're interested: It's pretty commonly known that international student-athletes in the NCAA face a lot more restrictions, like having to get a visa, being limited in the income they can make off name/image/likeness deals and so on. Despite or because of these factors, it's interesting to look into how international student involvement in women's basketball has been impacted over the years. Have more people become involved or been dissuaded from being involved, and how have those students contributed to women's basketball rosters? 
```{r}

international_students <- all_rosters |>
#The 2021-22 season is documented in two different ways in the data, so first we're renaming those values to fix that so the group by will be easier 
  mutate(season = case_when (season == "2021-2022" ~ "2021-22", 
    TRUE ~ season
  )) |>
#Then, since we want to only look at international students, we're filtering out all students from the USA and grouping by season to see the change in internationl numbers across seasons 
  filter(country_clean != "USA") |>
  group_by(season) |>
  summarise(
    international_count = n()
  ) 

  
#We know from our exploratory analysis that the 2020-21 season had 13075 players drafted, the 2021-22 roster had 13645 players drafted, the 2022-23 roster had 13803 players drafted, and the 2023-24 roster had 13907 players drafted. To show the number of international students as a percentage of the total players per season, we're going to mutate using these numbers 
international_students <- international_students |>
  mutate(total_season_players = case_when(
    season == "2020-21" ~ "13075",
    season == "2021-22" ~ "13645", 
    season == "2022-23" ~ "13803", 
    season == "2023-24" ~ "13907" 
    )) |>
  mutate(international_pct = as.numeric((international_count)/as.numeric(total_season_players))*100)

#To see what this trend looks like, we can make a visual 
international_students |>
  ggplot(aes(x=season, y=international_pct)) +
  geom_point() 
 
```

Answer: The share of international students that make up each season's NCAA women's basketball roster has grown fairly linearly from the 2020-21 season to the 2023-24 season, indicating an increase in the percentage of international students on each season's roster. 


4. What percentage of students on the 2023-24 roster are redshirts, and are players in a certain year more likely to redshirt?

Why we're interested: Since college athletes have five years to complete four seasons of their sport, they can redshirt in any year. We're interested in looking at what years are the most common 
```{r}

redshirt_23_24 <- roster_23_24 |>
  filter (redshirt == 1) |>
  group_by (year_clean) |>
  summarise (
    redshirt_count_by_year = n()
  ) |>
  arrange(desc(redshirt_count_by_year)) |>
  mutate(total_redshirts = sum(redshirt_count_by_year)) |>
#It might provide extra insight to see redshirts in a given year as a percentage of the total number of redshirts 
  mutate((redshirt_percentage = redshirt_count_by_year/total_redshirts)*100)

redshirt_23_24

#As stated in the limitations discussed above, we're unsure if the redshirt column indicates that a player redshirted in the 2023-24 season, or if it just means they redshirted at any point in their college career. We completed this question assuming that the data represents a redshirt IN the 2023-24 season (in other words, in the corresponding academic year listed in this same dataframe), but if the data was documented differently, the result for this could be wrong.

```
Answer: In the 2023-24 year, players were most likely to redshirt during their sophomore year. Junior year followed, and then freshman year. Senior year and a player's fifth year were the least common. In terms of the most and least popular options, about 31% more players redshirted during their sophomore year than their fifth year.


5. Are there differences in average height between the different divisions across all seasons? Also, is there a correlation between higher/lower average heights at different schools and how well those teams did? For the second question, we'll focus on just the 2023-24 season as the most recent example. 

Why we're interested: Height is often discussed as a major factor in someone's success as a basketball player, so it'll be interesting to look at whether taller or shorter average heights have driven players to more success (playing D1) or less success (playing D3), as well as if average height influenced how well individual teams did in the most recent season.

```{r}
#Use this codeblock to group by division, and then calculate average height for each division 
division_heights <- all_rosters |>
  filter(!is.na(division)) |>
  group_by(division) |>
  summarize(mean_height = mean(total_inches, na.rm = TRUE))

```

```{r}
#Use this codeblock to group by school, and then calculate average height at each school
#Then we'll look up rankings from the 2023-24 season and compare the tallest/shortest teams with these rankings to see if there are any relationships we can figure out. Since rankings are primarily focused on Division I, we'll filter out just Division I schools 

team_heights <- roster_23_24 |>
  filter(!is.na(total_inches)) |> 
  filter(division == "I") |>
  group_by(team) |>
  summarize(mean_height = mean(total_inches, na.rm = TRUE)) |>
arrange(desc(mean_height))
```

Answer: There are differences in average height across divisions. In Division III the average height (in inches) was 68.06 in. In Division II the mean was 69.15 in — slightly taller — and in Division I, the average at 70.55, marking the tallest average height. This shows a correlation between "stronger" players and taller heights. When zooming in on the 2023-24 season and looking at average height by teams in Division I, seven out of the 10 tallest teams on average were ranked in the top 25 of the division (https://www.espn.com/womens-college-basketball/rankings) in the NCAA for the season. In contrast, none of the shortest 10 teams on average were in the top 25 for the division. This shows that evidently, some of the tallest teams in the NCAA are also some of the best performing teams, whereas some of the shortest teams are not the best-performing. 
