---
title: "lab_12"
author: "Mitchell Hang"
date: "2023-05-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   tidytext and our usual libraries

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidytext)
library(tidyverse)
library(dplyr)
library(janitor)
library(lubridate)
library(rvest)
```

## Questions

**Q1.** You've been assigned to report a story about the leading reasons that Maryland attorneys get sanctioned by the state for misconduct. The state [publishes lists of sanctions](https://www.courts.state.md.us/attygrievance/sanctions) that contain a short text description about the situation. Load the CSV file in the data folder containing records from fiscal year 2011 onwards. Make a list of unique words from the text column, then following the example in the pre_lab, remove common "stop words" from that list and create a list of the top 10 words containing the percentage of occurrences each word represents. What's the leading word in that answer and, broadly, what do you think the top 10 words describe?

```{r}
#Load the data
md_attorney_sanctions <- read_csv("data/md_attorney_sanctions.csv")
```

```{r}
#Create list of unique words

unique_words <- md_attorney_sanctions |> select(text) |>
  unnest_tokens(word, text)
View(unique_words)

```

```{r}
#Load stop words
data("stop_words")

```

```{r}
#Remove stop words from unique words list
unique_words |>
  anti_join(stop_words) |>
  group_by(word) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)

```

**A1.**: Failing is the leading word in this list. It seems like the top 10 words are describing sanctions where attorneys failed to do something, did things without consent, or were involved in fraudulent things related to funds/accounts. 

------------------------------------------------------------------------

**Q2.** Let's move beyond single words to phrases. Make a list of the top 10 three-word phrases, called trigrams, based on the example from the pre_lab (you'll need to modify the example code to do this). What's the top trigram and how often does it appear? What does that phrase mean in legal terms?

```{r}
# Check the task 12 on Pre-lab 12 to see if you can use that code

md_attorney_sanctions |>
  unnest_tokens(trigram, text, token = "ngrams", n = 3) |>
  separate(trigram, c("word1", "word2", "word3"), sep = " ") |>
  filter(!word1 %in% stop_words$word) |>
  filter(!word2 %in% stop_words$word) |> 
  filter(!word3 %in% stop_words$word) |> 
  mutate(trigram = paste(word1, word2, word3, sep=" ")) |>
  group_by(trigram) |>
  tally(sort=TRUE) |>
  mutate(percent = (n/sum(n))*100) |>
  top_n(10)

```

**A2.**: The top trigram is "attorney trust account," and it shows up 343 times in the sanctions, making up almost 6 percent of the total words in the sanctions. An attorney trust account is a particular type of account where client funds are kept until it's time to withdraw them, to ensure that they're not mixed up with other sources of law firm funding. 

------------------------------------------------------------------------

**Q3.** Let's drop back down to more traditional text analysis - take the top trigram from Q2 and write code to see how many times it occurs in the text column in each fiscal year. What do you think the answer produced by your code suggests? What else could you do to try and clarify the most important reasons attorneys get sanctioned?

```{r}
fiscal_year_sanctions <- md_attorney_sanctions |>
  filter(str_detect(text, "attorney trust account")) |>
  group_by(fiscal_year) |>
  summarise (
    phrase_count = n()
  ) 

fiscal_year_sanctions

```

**A3.**: This code suggests that sanctions for misuse of attorney trust accounts have jumped in more recent years — for instance, between 2016 and 2017, we see them more than triple. There are some dips, like in 2018, but for the most part, these types of sanctions have been steadily increasing over the years. Some other things we could do: group the other important trigrams (and bigrams) by fiscal year to see if there is a certain one that shows up more commonly, and then looking at the text tself to search for that term and understand the context around it. We could also do some background research about common attorney sanctions, and then narrow down the terms we look for in this particular dataset based on that background knowledge.