---
title: "lab_03"
author: "Derek Willis"
date: "2023-02-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
md_agency_payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv")


# Load in the Maryland grants and loan data
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv")

cleaned_md_agency_payments <- md_agency_payments |> 
  clean_names()

cleaned_md_grants_loans <- md_grants_loans |>
  clean_names()
  
```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1: I found 66 records of payments made to the county school system. 

```{r}
pg_schools <- 
cleaned_md_agency_payments |> 
filter(str_detect(vendor_name, "SCHOOL") & str_detect(vendor_name, "PRINCE") | str_detect(vendor_name, "PG") & str_detect(vendor_name, "SCHOOL")) 
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? What do you notice about the years listed, and what questions do you have about that data?

A2: Based on all the records I found in the previous question, 2021 had the largest total amount, at $1,470,133,533. I noticed that the years listed span from 2009 to 2021, but that some years like 2011, 2012 and 2016 don't have any payments listed to the school system. First, I wonder why there are no payments listed during these years. Second, I wonder how the spending data varied from year to year — are there different agencies who contributed more or less in certain years?

```{r}
pg_schools |> 
  group_by(fiscal_year) |>
  summarise (
    fiscal_year_amounts = sum(amount) 
  ) |>
  arrange(desc(fiscal_year_amounts))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. We will need to remove any possible duplicates to ensure that we only have one row per vendor and zip code. For instance, Whiting Turner Contracting Co and Whiting Turner Contracting Company are two separate rows that should be combined. Grim & Parker Architects and Grimm + Parker Architects should probably also be combined. Ballinger and The Ballinger Co should also be combined. To clean this data, I might first use the get_dupes() function and then distinct() to get rid of rows that are exactly the same and repeated. To deal with the rows that have slight variation but still need to be combined, I would write a line of code to merge the lines of code that need to be combined, and then a line of code to rename the vendor_name for those to a standardized name. 

```{r}
cleaned_md_agency_payments |>
  filter(str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND")) |>
  group_by(vendor_name) |>
  summarise (
    vendor_total = sum(amount)
  ) |>
  arrange(desc(vendor_total))

```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: Stem cell research funding has, on average, dropped over time, with fluctuations seeing certain years receive slightly more funding. 

```{r}
cleaned_md_grants_loans |> 
  mutate(description = str_to_upper(description)) |>
  filter(str_detect(description, "STEM CELL")) |>
  group_by (fiscal_year) |>
  summarise (
    stem_cell_total = sum(amount)
  ) |>
  arrange(fiscal_year)
```
