---
title: "lab_06"
author: "derek willis"
date: "2022-03-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# load your libraries
library(tidyverse)
library(janitor)
library(lubridate)

```

## Getting Started

This week we'll be combining and joining data to make it more useful and to ask some questions about it. We'll use some of our 911 overdose call data to do this. The first thing we want to do is to combine multiple counties' data into a single dataframe so that we can ask some questions. First, let's combine data from Cecil, Carroll and Allegany counties into a new dataframe.

*Before* you combine them you'll need to clean up some of the column names to make the data easier to work with - make sure you have loaded the library to do that. You also need to make sure that each column has the same name and datatype (you can check the datatypes using `glimpse`). If any of them does not, you need to fix that.

```{r}
# load and combine the call data from those three counties
cecil_911 <- read_csv("data/cecil_911.csv") |>
  glimpse() 

carroll_911 <- read_csv("data/carroll_911.csv") |>
  glimpse() |> clean_names()


allegany_911 <- read_csv("data/allegany_911.csv") |>
  glimpse()

counties_911_combined <- bind_rows(list(cecil_911, carroll_911, allegany_911))

```

Then we'll join some demographic data with Baltimore City 911 calls and ask some questions of that. For that, you'll need to load the Baltimore City 911 data and the ZCTA demographic data, and join them based on the zip code. Remember, the zip codes in both dataframes must be the same datatype (and should be characters). You should start with your Baltimore 911 dataframe in creating your join.

```{r}
# load and join the Baltimore-related data
baltimore_911 <- read_csv("data/baltimore_911.csv") |>
  mutate(zip_code = as.character(zip_code))

demographic_data <- read_csv("data/maryland_zcta.csv") |> clean_names() |>
  mutate(zip_code = as.character(zcta5n))

baltimore_and_demographic_joined <- baltimore_911 |> left_join(demographic_data, join_by("zip_code"))


```

## Answer questions

### Q1. Let's start with our data from Allegany, Cecil and Carroll counties. Write code to generate a dataframe that shows the total number of calls for each county. What's the order of counties from most calls to least?

A1. Carroll, Cecil, Allegany

```{r}
counties_911_combined |>
  group_by(county) |>
  summarise (
    county_total_calls = n()
  ) |>
  arrange(desc(county_total_calls))
```

### Q2. In your combined dataframe that you used in A1, add a column for the month of each call, then write code to show the total number of calls per county and month. Are there any outliers or noteworthy results? Describe the general pattern of the results.

Then calculate the most calls per county and location. Which location had the most calls - where and what is that location?

A2.It looks like the most calls across counties happen in the warmer months, since the top results range from May to September. Cecil County in August is definitely somewhat of an outlier, with 55 calls. Allegany County in January and November is also an outlier, with less than 20 calls each in both months. When looking at the data by month, Allegany County in June is an outlier, with significantly fewer calls than the other two counties in the same month. But in September, it had significantly more calls than the other two counties did. Generally, the data stays relatively steady from month to month, with slight increases in the number of calls across all the counties during the summer months. 

314 Grove Neck Road in Earleville, Maryland had the most calls, which is in Cecil County. It looks like this is an addiction treatment center that is part of the Recovery Centers of America network.

```{r}
counties_911_combined |>
  mutate(month= month(date)) |>
  group_by (county, month) |>
  summarise (
    county_and_month_calls = n()
  ) |>
  arrange(month)

counties_911_combined |>
  mutate(month= month(date)) |>
  group_by (county, location) |>
  summarise (
    county_and_location_calls = n()
  ) |>
  arrange(desc(county_and_location_calls))

```

### Q3. Let's turn to the Baltimore City data. You're interested in exploring calls in zip codes with a high percentage of children under 18. Write code to isolate calls in zip codes where the percentage of under 18 population is at least 75% and then show the zip code and population and how many calls occurred in each of those zip codes.

Where are those zip codes in the city?

Add a column to your code that calculates the number of calls per 1,000 people for those zip codes. To calculate a per capita rate, you might find this short guide handy: [https://observablehq.com/\@palewire/per-capita-calculator](https://observablehq.com/@palewire/per-capita-calculator){.uri}.

Which zip code has the highest rate of calls per 1,000 people? Find the neighborhoods in that zip code that are listed in the data - you can use group_by or distinct to do this. What are some of those neighborhoods?

A3.The five zip codes with a high percentage of under 18 population all are concentrated around the center of Baltimore/the downtown area. 21217 has the highest rate of calls per 1,000 people. The neighborhoods in the 21217 zip code include Penn North, Harlem Park, Druid Heights, Bolton Hill and Reservoir Hill.

```{r}
baltimore_and_demographic_joined |> 
  filter (pct_under_18 >= 75) |>
  group_by(zip_code, population) |>
  summarise (
    calls_per_zip_code = n()
  ) |>
  arrange(desc(calls_per_zip_code))

baltimore_and_demographic_joined |>
  filter(pct_under_18 >= 75) |>
  group_by(zip_code, population) |>
  summarise (
    calls_per_zip_code = n()
  ) |>
  mutate(calls_per_capita = (calls_per_zip_code/population)*1000) |>
  arrange(desc(calls_per_capita))

baltimore_and_demographic_joined |>
  filter(zip_code== "21217") |>
  select(zip_code, neighborhood) |>
  distinct()
  
```

### Q4. Choose your own adventure: I want you to come up with a question you could ask of the Baltimore City data that combines both the calls and the demographic information. You must use both demographics and the overdose call data in your work.

A4: I wanted to see if the percentage of homeowners in a zip code impacts the number of overdose calls there are in that neighborhood. Coding this out showed me that, as expected, the most calls happened in zip codes where there are a higher percentage of renters.

```{r}
baltimore_and_demographic_joined |>
  group_by(zip_code, pct_rent) |>
  summarise (
    number_of_calls = n()
  ) |>
  arrange(desc(number_of_calls))
```
