---
title: "lab_04"
author: "Derek Willis"
date: "2023-01-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). Read the questions carefully and attempt to do all of what they ask. Make a list if it helps.

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)
library(tidyverse)
library(lubridate)
library(janitor)

```

For this lab, we want to investigate spending by Maryland state and local candidates during the 2022 election. For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis. The function to export data from RStudio is called `write_csv()` and it takes two arguments: the name of the dataframe and the path of the file to write it to, which is in quotes. Like `read_csv()`, but the variable name comes inside. Here's an example, in which you'd replace the dataframe and the path values.

```{r}
write_csv(dataframe, "folder/file.csv")
```

## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table, clean the column names and convert columns with dates into actual dates.
md_expenses <- read_csv("data/maryland_expenses.csv") |>
clean_names() |>
mutate (expenditure_date=mdy(expenditure_date))
```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". Write code to isolate Media expenses where the name of the payee is not NA. Then, using the `write_csv` function, make a CSV file of those media expenses in your data folder. Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing your work on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: <https://about.fb.com/news/2021/10/facebook-company-is-now-meta/>).

Then, when you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving that file a different name* and read it back into RStudio as a new dataframe.

Then, using that `payee_clean` column, write code to find all of the spending where Facebook is the payee and calculate the total amount of spending for each committee and present the results showing the committee that spent the most first. Which committee is that?

**A1.**: Friends of Ashwani Jain

```{r}
# One step at a time. Find the category column in the data that has values for different kinds of spending.
#The column is expense_category.

# One of them is "Media". Write code to isolate Media expenses where the name of the payee is not NA. They key here is using (!is.na(payee_name)) within a function

media_expenses <- md_expenses |>
filter(expense_category == "Media") |>
filter(!is.na(payee_name))

```

```{r}
# Export this table as a CSV

write_csv(media_expenses, "data/media_expenses.csv")
 

# Clean up work in OpenRefine

```

```{r}
# read the cleaned CSV 

media_expenses_clean <- read_csv("data/media-expenses-clean.csv") 

# calculate top Facebook spenders

facebook_spending <- media_expenses_clean |>
filter(payee_clean == "Facebook") |>
group_by(committee_name) |>
  summarise(
    total_facebook = sum(amount)
  ) |>
arrange(desc(total_facebook))

```

**Q2.** Let's do some research on the top committees from question 1. Look up the top 10 spenders: if they are candidate committees, what offices were they running for? If they were not candidates, what issues do they seem to care about? Write some sentences answering those questions, then go the Facebook ad library ([https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\\[direction\\]=desc&sort_data\\[mode\\]=relevancy_monthly_grouped&media_type=all](https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data%5Bdirection%5D=desc&sort_data%5Bmode%5D=relevancy_monthly_grouped&media_type=all){.uri}) and look for ads from the top spender. What do you notice about those ads?

**A2.** Most of the top 10 spenders are candidate committees. Ashwani Jain ran for U.S. Congress to represent Maryland's 6th district. Peter Franchot and Monique Anderson-Walker ran for governor and lieutenant governor of Maryland. Another top spender was Peter Franchot's individual committee, which likely spent on his campaign before he partnered with Anderson-Walker and created a new candidate committee. Billy Boniface ran for Harford County Executive. Jon Baron ran for governor of Maryland. WFP National Pac is the Working Families Party, a third party that advocates for workers. John King and  Michelle Siri ran for governor and lieutenant governor. John King's individual candidate committee was also from when he ran for governor. Amanda Fiedler ran for Anne Arundel County Council to represent District 5. Firefighters of Frederick County for Collective Bargaining is a committee of firefighters in Frederick County advocating for collective bargaining rights.

When looking at ads paid for by Friends of Ashwani Jain, I noticed that they were pretty frequent, with at least one a day, if not multiple. The ads also included some sort of multimedia component, like a campaign poster or a photo of Jain.

**Q3.** Given the list of top spenders in A1 and the primary results for governor (<https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html>), who is missing from that A1 list? Did those campaigns run Facebook ads, according to Facebook's ad library? Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees)?

**A3.** Wes Moore and Dan Cox are missing from the list of top spenders — they each won their respective primaries but didn't spend that much on Facebook ads, evidently. Based on Facebook's ad library, Dan Cox's campaign did not run Facebook ads. Wes Moore's campaign did, but a lot of those ads have been removed because they didn't align with Facebook's policy for political ads. The biggest recipients of spending by Moore's committees are SKDK, a public affairs consulting firm, and Foglamp, a digital strategy company. Other top recipients are Apollo Artistry, a digital creative agency, and Shauna Henson, a brand strategist.

```{r}
dem_spending <- media_expenses_clean |>
  filter(str_detect(committee_name, "Moore")) |>
  group_by (payee_name) |>
  summarise (
    top_dem_recipient = sum(amount) 
  ) |>
  arrange(desc(top_dem_recipient))

```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**A4.** This list does not really align with the list of top five Facebook spenders. The committees that spent the most on online advertising seem to have directed most of their money elsewhere, because none of them appear in the Facebook spenders list. One exception is Jon Baron, whose candidate committee Jon Baron for Maryland appeared in the top Facebook spenders, and whose joint candidate committee with his running mate (Baron-Williams Slate For Maryland) appears in the top online advertising spenders. Of the top five Facebook spenders, Ashwani Jain's committee seems to have spent all its online advertising dollars on Facebook.So does Franchot and Anderson-Walker's candidate committee. 

```{r}
online_ad_spending <- media_expenses_clean |>
  filter(expense_purpose == "Online Advertising") |>
  group_by (committee_name) |>
  summarise (
    online_amount = sum(amount)
  ) |>
  arrange(desc(online_amount))

```

**Q5** You've been asked to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**A5**: We know who the top spenders are, and that the top Facebook spender spent almost $5,000 more than any other committee on Facebook advertising. I think this is probably one of the best story ideas, and it would be good if we also analyzed the content of the top spenders' ads and looked for trends in how they advertised on Facebook in particular. We also know which candidate committees spent the entirety of their advertising money on Facebook, and which ones didn't. Right now, without extra research, we don't know the success rates of various candidates compared to how much they spent on Facebook. I think it would be a good story idea to examine what percentage of the vote each candidate got in their respective race and see if there are any links between that and their candidate committee's Facebook spending amount (or just whether they won). 
