---
title: "lab_10"
author: "Derek Willis"
date: "2023-04-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Our usual libraries, plus some of the map-specific ones from the pre_lab. You'll also be fetching Census data in Q3. Read the questions carefully!

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
options(scipen=999)
# Load libraries
library(tidyverse)
library(janitor)
library(tidycensus)
library(tigris)
library(sf)
```

Let's explore Maryland elections, power outages and electric vehicle data and make some maps!

## Questions

**Q1.** Maryland voters made history in 2022 by electing the state's first two African-American statewide office holders: Wes Moore as governor and Anthony Brown as attorney general. You want to compare the two's electoral performances. Using the `md_statewide_county.csv` file in the data folder, calculate the total number of votes for both the governor's race and the attorney general's race, then percentages for both Moore and Brown *based on the total of their respective contests*. Then calculate the difference between their percentages.

Get the geometry for Maryland's counties using tigris and, joining that dataframe to your results dataframe, make a county-level map of the differences between Moore and Brown's percentages. What is the story here, and what is the county with the most interesting results?

**A1.** The story is that five counties in Maryland, Moore won significantly more of the vote than Brown did. Reporting on this would focus on voter attitudes in those counties and determine what pushed some people to presumably vote for Moore in one race, but not vote for Brown in the other. Kent County is the most interesting because it shows the biggest Moore-Brown vote difference.

```{r}
#Calculate percentages for both candidates and the difference between them per county
elections <- read_csv("data/md_statewide_county.csv") |>
  clean_names() |>
  mutate(gov_votes = cox+moore+lashar+wallace+harding+sefcik+gov_writein) |>
  mutate(ag_votes = peroutka+brown+ag_writein) |>
  mutate(moore_pct = (moore/gov_votes) *100) |>
  mutate(brown_pct = (brown/ag_votes) *100) |>
#This will show how much more of the vote Moore got in a county than Brown did in that same county. If it comes out as a negative number, that means Brown got a higher percentage of the vote in that county than Moore did in his race.
  mutate(moore_brown_difference = moore_pct - brown_pct)
```

```{r}
#Take county geometry from tigris
counties <- counties()
glimpse (counties)


#Filter out MD counties
md_counties <- counties |>
filter(STATEFP == "24") |>
  clean_names()

elections <- elections |>
  mutate(geoid= as.character(geoid))
```

```{r}
#Merge geometries with voting data
 elections_and_counties <- md_counties |> left_join(elections, join_by(geoid==geoid))


```

```{r}
#Create the map

ggplot() +
  geom_sf(data=elections_and_counties, aes(fill=moore_brown_difference)) +
  scale_fill_viridis_b(option="magma")

```

**Q2.** Make a county-level map to explore the number of customers impacted by power outages in Maryland. Load the `county_outages_march2023.csv` file in the data folder and calculate a rate of customers with outages per 10,000 total customers for each county. Then join that dataframe to your Maryland counties dataframe you made in Q1 and create a county-level map of the results; you may need to tweak the fill scale to make the distinctions clearer. What's the potential story here? Given access to other months' outage data, what would your next steps be in reporting a story?

**A2.**: The story here is that Talbot County has the highest rate of outages per 10,000 customers, and Frederick County has the lowest rate. If I had access to other months' data, I'd plot those for all the counties and see if things have changed, or if those two counties have consistently been at either end of the power outage scale.

```{r}
#Read data and calculate rate of outages per customer
county_outages <- read_csv("data/county_outages_march2023.csv") |>
  
  mutate(rate=(outage_customers/total_customers*10000))
  

```

```{r}
#Merge geometries with outage data

#Fixing repeat Baltimore 
md_counties_fixed <- md_counties |>
  mutate(name = case_when(
    namelsad == "Baltimore city" ~ "Baltimore City", 
    .default = name)) 

outages_and_counties <- md_counties_fixed |> left_join(county_outages, join_by(name==county))

```

```{r}
#Map it out
ggplot() +
  geom_sf(data=outages_and_counties, aes(fill=rate)) +
  scale_fill_viridis_c(option="magma", trans="log")

```

**Q3.** You've been assigned a story about the use of electric vehicles in Maryland and you want to make a map of the prevalence of electric vehicles in each county. Load the `ev_2020_2023.csv` file, which has the number of electric and hybrid vehicles per county from July 2020 and March 2023 and calculate a percentage change for each county (remember, that's `new-old/old`). Using the Maryland counties dataframe you made for Q1, make a county map showing the percentage change.What counties stand out to you? What questions do you have about the changes you're seeing?

Next, use tidycensus to retrieve county population data for Maryland from 2021 (you did this in the pre_lab) and join it to your vehicles dataframe. HINT: you may not need to join it on the county name, and you should already have the geometry!

Then, calculate a per capita rate of electric & hybrid vehicles using the March 2023 figure and the population estimate. You will need to choose the "per capita" value.

Finally, make a second county map showing the per capita rate. What counties stand out on this map? How is it different from the first one you made?

**A3.**: Caroline County stands out to me because it's had the biggest growth in EV usage. Dorchester and Washington counties also stand out because they've had the lowest rate of growth. I want to know what's spurred these changes, or lack thereof, and if there have been any county-level policies in these places regarding EVs that have impacted the changes.

```{r}
#Read data and calculate change of EV vehicles
ev_usage <- read_csv("data/ev_2020_2023.csv") |>
  mutate(percent_change = (march_2023 - july_2020)/july_2020) 

```

```{r}
#Merge geometries with EV data
ev_and_counties <- md_counties_fixed |> left_join(ev_usage, join_by(name==county)) |>
  arrange(desc(percent_change))

```

```{r}
#Map out the change
ggplot() +
  geom_sf(data=ev_and_counties, aes(fill=percent_change)) +
  scale_fill_viridis_c(option="magma", trans="log")


```

```{r}
#Grab population data from Census
census_counties <- get_acs(geography = "county",
              variables = c(population = "B01001_001"),
              year = 2021,
              state = "MD",
              geometry = TRUE) |>
  clean_names()
```

```{r}
#Calculate per capita rate for March 2023

#Joining census data with ev data 
evs_and_population <- ev_and_counties |> inner_join(census_counties, by=("geoid"))
```

```{r}
#Map out the per capita rate

```
